\documentclass{article}
\usepackage{graphicx} % Required for inserting images

\title{Improving variable selection properties with transfer learning and data integration}
\author{Paul Rognon}
\date{October 2025}

\begin{document}

\maketitle

Transfer learning transferring the knowledge acquired in a source dataset to a 
target dataset. We study here how transferring the set of variables learned to be associated with the source outcome can ease variable selection in the target. 
More generally, we study how variable selection properties can be improved by leveraging side information on the likelihood of variables to be truly associated 
with the outcome. We introduce a family of penalized linear regression methods where the penalties depend on the side information and are motivated by connections 
to Bayesian variable selection. We develop a careful analysis of the fundamental limits on consistent variable selection in linear regression and show that these 
can be pushed with side information and transfer learning. We precisely quantify the gains that are achievable for ideal penalties set by an oracle. We subsequently 
propose computationally efficient algorithms for the incorporation of side information and transfer learning that are inspired by empirical Bayes techniques and do 
not require an oracle. We show these algorithms recover most of the gains achievable by oracle penalties.  

\section{Introduction}

frequentist guarantees of Bayesian methods
transfer learning amounts to learning for the target data a good prior from the source data.  

\end{document}
